Improving the prospects for multimessenger astronomy with \earlywarning\ detection of compact binary coalescence.

### 2011-05-17 introduction.tex
Added a new paragraph on optical flashes. Still needs some work.
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@754 ccdd05d5-56c1-461a-bf98-85f6876523b4

As a compact binary system loses energy to gravitational waves, its orbital separation
decreases. This causes a run-away inspiral with the gravitational-wave
amplitude and frequency increasing until the system eventually merges near the
innermost stable circular orbit (\ISCO).  If a neutron star is involved it may become tidally disrupted near the merger.
This disrupted matter can fuel a bright electromagnetic
counterpart in the system's final moments as a binary~\cite{shibata:2007}.
In order to observe the prompt and
potentially most intense electromagnetic emission, telescopes must point during the seconds surrounding the
merger.

\editorial{Citation needed for LOOC-UP} To make this possible, the gravitational-wave community initiated a project to send alerts when
potential gravitational-wave transients are observed.  In October 2010, \textsc{ligo} completed its sixth science run (S6) and Virgo
completed its third science run (VSR3).  While both \textsc{ligo} detectors and Virgo were
operating, several all-sky detection pipelines operated in a low-latency
configuration, namely \textsc{mbta}, ihope, Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}. \editorial{Get references for these low-latency pipelines.} The S6 analysis achieved latencies of 30--60 minutes, which were
dominated by a human vetting process. Candidates were sent for
electromagnetic followup to several telescopes; Swift,
\textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008, HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved the best gravitational-wave trigger
generation latencies of 2--5 minutes.  We assume that
in the advanced detector era the vetting process will be automated, so current
trigger generation and telescope actuation would then dominate latency.

To this end, we have the ambition of reporting candidates not minutes \emph{after} the merger, but seconds \emph{before}.  By looking for threshold crossings before the signal leaves the detection band, it is possible to trade some signal to noise ratio (\textsc{snr}) for latency.  Figure \ref{fig:earlywarning} shows projected early trigger rates for NS--NS binaries in Advanced \LIGO\
assuming the event rate predictions in \cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $T_{bc}$ seconds before coalescence.  The solid line is the most
likely yearly rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the
shaded region is the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources may be detected within 10
seconds of merger and $\sim1$ sources maybe detected within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early
detection but at the cost of fewer total events observed above SNR 8.}
\end{figure}

Predictive detection of CBCs will require striking a balance between latency and throughput.  CBC searches consist of banks of matched filters, or cross-correlations between the data stream and a bank of nominal ``template'' signals.  There are many different implementations of matched filters, but most have high throughput at the cost of high latency, or low latency at the cost of low throughput.  The former are epitomized by the overlap-save algorithm for \textsc{fft} convolution, currently the preferred method in gravitational wave searches.  The most obvious example of the latter is the time domain (\textsc{td}) convolution, which is has no latency at all.  However, its computational complexity is quadratic in the length of the templates, so it is prohibitively expensive for long templates.

>>>>

As a compact binary system loses energy to gravitational waves, its orbital separation
decreases. This causes a run-away inspiral with the gravitational-wave
amplitude and frequency increasing until the system eventually merges near the
innermost stable circular orbit (\ISCO).  If a neutron star is involved it may become tidally disrupted near the merger.
This disrupted matter can fuel a bright electromagnetic
counterpart in the system's final moments as a binary~\cite{shibata:2007}.

Prompt electromagnetic emission can arise as shells of relativistically
outflowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be the mechanism of short gamma-ray
bursts~\cite{nakar07}. This prompt emission is a probe into the extreme initial
conditions of the outflow. A very bright optical flash should accompany the
x-rays and gamma-rays~\cite{Sari99}. \editorial{Should mention that optical flashes have been seen for a handful of long GRBs by very fast slews. TAROT has seen many within ~20 seconds. \cite{Klotz09}} In contrast, an
afterglow is produced when the outflow impacts the interstellar medium.
Afterglows represent lower Lorentz factors and the conditions of the
environment far outside of the central engine, and are relatively insensitive
to initial conditions. In order to observe the most intense electromagnetic
emission and provide an unbroken chain of observations between the progenitor
system and the expanding fireball, telescopes must point during the seconds
surrounding the merger.

\editorial{Citation needed for LOOC-UP} The gravitational-wave community initiated a project to send alerts when
potential gravitational-wave transients are observed.  In October 2010, \textsc{ligo} completed its sixth science run (S6) and Virgo
completed its third science run (VSR3).  While both \textsc{ligo} detectors and Virgo were
operating, several all-sky detection pipelines operated in a low-latency
configuration, namely \textsc{mbta}, ihope, Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}. \editorial{Get references for these low-latency pipelines.} The S6 analysis achieved latencies of 30--60 minutes, which were
dominated by a human vetting process. Candidates were sent for
electromagnetic followup to several telescopes; Swift,
\textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008, HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved the best gravitational-wave trigger
generation latencies of 2--5 minutes.  We assume that
in the advanced detector era the vetting process will be automated, so current
trigger generation and telescope actuation would then dominate latency.

To this end, we have the ambition of reporting candidates not minutes \emph{after} the merger, but seconds \emph{before}.  By looking for threshold crossings before the signal leaves the detection band, it is possible to trade some signal to noise ratio (\textsc{snr}) for latency.  Figure \ref{fig:earlywarning} shows projected early trigger rates for NS--NS binaries in Advanced \LIGO\
assuming the event rate predictions in \cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $T_\mathrm{bc}$ seconds before coalescence.  The solid line is the most
likely yearly rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the
shaded region is the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources may be detected within 10
seconds of merger and $\sim1$ sources maybe detected within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early
detection but at the cost of fewer total events observed above SNR 8.}
\end{figure}

Predictive detection of \CBC{}s will require striking a balance between latency and throughput.  \CBC{} searches consist of banks of matched filters, or cross-correlations between the data stream and a bank of nominal ``template'' signals.  There are many different implementations of matched filters, but most have high throughput at the cost of high latency, or low latency at the cost of low throughput.  The former are epitomized by the overlap-save algorithm for \textsc{fft} convolution, currently the preferred method in gravitational wave searches.  The most obvious example of the latter is the time domain (\textsc{td}) convolution, which is has no latency at all.  However, its computational complexity is quadratic in the length of the templates, so it is prohibitively expensive for long templates.
